---
title: "CEPM WJ Energy Index Transformation"
author: "Tom Lombardi"
date: "February 9, 2016"
output: html_document
---

#W&J Energy Index Database Construction
This website documents the methods for building the CEPM database that 
tracks the W&J Energy Index for the Center fo Energy Policy and Management.
The RK Mellon Grant supported this work.

##Understanding the Data
Drs. Leslie and Robert Dunn developed the W&J Energy Index to help 
economists understand the dependence of the United States on foreign 
sources of energy. The researchers prepare their data in two 
spreadsheets representing the National and Regional/State level index values, 
weights and components. Spreadsheets are extremely convenient for ad hoc 
analysis of data sets, but pose problems when presented on the world wide 
web. The R code documented below transforms the spreadsheet data into a set 
of SQL statements for the MySQL environment.

##Entity-Relationship Diagram

##DDL for Base Tables

##Transformations for Measurements
###Required Packages
The code below requires the installation of two packages.
```
install.packages("XLConnect")
install.packages("reshape2")
```
`XLConnect` provides easy access to the data in Excel spreadsheets.
`reshape2` offers methods for reshaping data into useful formats

###Processing the National-level Data
The code below connects to the `National.xlsx` spreadsheet to process
the national-level indices. After the entire workbook is loaded, five 
different sheets are stored in data frames.
```{r}
library("XLConnect")
wb <- loadWorkbook("national.xlsx")
i1 <- readWorksheet(wb, sheet=1)
i2 <- readWorksheet(wb, sheet=2)
w <- readWorksheet(wb, sheet=3, startRow=2)
ps <- readWorksheet(wb, sheet=4, startRow=2)
pse <- readWorksheet(wb, sheet=5,startRow=2)
```

###Processing Sheet 1
Sheet 1 contains the W&J Energy Index values.
The code below performs the following steps.

* Adds columns to the data frame to store the region, source and component codes
required by the database.

* Reorganizes and renames the columns to match the database design.

* Loops over each row in the data frame to produce the required SQL insert
statements.

* Writing the data to a file called `export_index1.sql`.

```{r}
i1$reg_id <- 1  
i1$src_id <- 12 
i1$comp_id <- 1 

expi1 <- data.frame(cbind(i1$Year,i1$reg_id,i1$src_id,i1$comp_id,i1$W.J.Energy.Index))
names(expi1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(expi1)){
  text <- paste(text, "insert into tbl_values values (",
                expi1[i,1],",",
                expi1[i,2],",",
                expi1[i,3],",",
                expi1[i,4],",",
                expi1[i,5],");\n",sep="")
}
fileConn<-file("export_index1.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 2
The steps for processing Sheet 2 are the same as those for Sheet 1.
```{r}
text <- NULL
i2$reg_id <- 1  
i2$src_id <- 12
i2$comp_id <- 2

expi2 <- data.frame(cbind(i2$Year,i2$reg_id,i2$src_id,i2$comp_id,
         i2$W.J.Energy.Index..with.consideration.of.exports.))
names(expi2) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(expi2)){
  text <- paste(text, "insert into tbl_values values (",
                expi2[i,1],",",
                expi2[i,2],",",
                expi2[i,3],",",
                expi2[i,4],",",
                expi2[i,5],");\n",sep="")
}
fileConn<-file("export_index2.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 3
Sheets 3, 4 and 5 have a more complex structure than previously discussed 
Sheets. These data are organized in a wide format and we must reorganize them
into a long format. The `reshape2` package makes this straight-forward.

* Load the `reshape2` package.

* Use `melt()` to convert the format of the data from a wide to a long format.

* Add additional columns for region, component and source data.

* Set the value of the src_id to the appropriate key value in the database.

* Reorganize and rename data frame to match the database design.

* Loop over the data frame to construct the necessary SQL insert statements.

* Write the SQL to a file called `export_weights.sql`

```{r}
library(reshape2)
ww <- melt(w,id.vars="Year")
ww$reg_id <- 1
ww$comp_id <- 3 
ww$src_id <- 1
ww[ww$variable=="Natural.Gas",]$src_id <- 2
ww[ww$variable=="Petroleum",]$src_id <- 3
ww[ww$variable=="Nuclear",]$src_id <- 4
ww[ww$variable=="Hydro",]$src_id <- 5
ww[ww$variable=="Geothermal",]$src_id <- 6
ww[ww$variable=="Solar.PV",]$src_id <- 7
ww[ww$variable=="Wind",]$src_id <- 8
ww[ww$variable=="Wood.Waste",]$src_id <- 9
ww[ww$variable=="Fuel.Ethanol",]$src_id <- 10
ww[ww$variable=="Biodiesel",]$src_id <- 11

expw <- data.frame(cbind(ww$Year,ww$reg_id,ww$src_id,ww$comp_id,ww$value))
names(expw) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(expw)){
  text <- paste(text, "insert into tbl_values values (",
                expw[i,1],",",
                expw[i,2],",",
                expw[i,3],",",
                expw[i,4],",",
                expw[i,5],");\n",sep="")
}
fileConn<-file("export_weights.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Comparing wide and long format data
**Wide Format:Before**
```{r}
head(w)
```
**Long Format:After**
```{r}
head(ww)
```

###Process Sheet 4
```{r}
psexp <- melt(ps,id.vars="Year")
psexp$reg_id <- 1 
psexp$comp_id <- 4 
psexp$src_id <- 1
psexp[psexp$variable=="Natural.Gas",]$src_id <- 2
psexp[psexp$variable=="Petroleum",]$src_id <- 3
psexp[psexp$variable=="Nuclear",]$src_id <- 4
psexp[psexp$variable=="Hydro",]$src_id <- 5
psexp[psexp$variable=="Geothermal",]$src_id <- 6
psexp[psexp$variable=="Solar.PV",]$src_id <- 7
psexp[psexp$variable=="Wind",]$src_id <- 8
psexp[psexp$variable=="Wood.Waste",]$src_id <- 9
psexp[psexp$variable=="Fuel.Ethanol",]$src_id <- 10
psexp[psexp$variable=="Biodiesel",]$src_id <- 11

expps <- data.frame(cbind(psexp$Year,psexp$reg_id,psexp$src_id,psexp$comp_id,psexp$value))
names(expps) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(expps)){
  text <- paste(text, "insert into tbl_values values (",
                expps[i,1],",",
                expps[i,2],",",
                expps[i,3],",",
                expps[i,4],",",
                expps[i,5],");\n",sep="")
}
fileConn<-file("export_index1_sources.sql")
writeLines(text, fileConn)
close(fileConn)
```
###Process Sheet 5
```{r}
pseexp <- melt(pse,id.vars="Year")
pseexp$reg_id <- 1 
pseexp$comp_id <- 5 

pseexp$src_id <- 1
pseexp[pseexp$variable=="Natural.Gas",]$src_id <- 2
pseexp[pseexp$variable=="Petroleum",]$src_id <- 3
pseexp[pseexp$variable=="Nuclear",]$src_id <- 4
pseexp[pseexp$variable=="Hydro",]$src_id <- 5
pseexp[pseexp$variable=="Geothermal",]$src_id <- 6
pseexp[pseexp$variable=="Solar.PV",]$src_id <- 7
pseexp[pseexp$variable=="Wind",]$src_id <- 8
pseexp[pseexp$variable=="Wood.Waste",]$src_id <- 9
pseexp[pseexp$variable=="Fuel.Ethanol",]$src_id <- 10
pseexp[pseexp$variable=="Biodiesel",]$src_id <- 11

exppse <- data.frame(cbind(pseexp$Year,pseexp$reg_id,pseexp$src_id,pseexp$comp_id,pseexp$value))
names(exppse) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(exppse)){
  text <- paste(text, "insert into tbl_values values (",
                exppse[i,1],",",
                exppse[i,2],",",
                exppse[i,3],",",
                exppse[i,4],",",
                exppse[i,5],");\n",sep="")
}
fileConn<-file("export_index2_sources.sql")
writeLines(text, fileConn)
close(fileConn)

```

###Processing Regional and State Data
The regional- and state-leve data only include index1 data.

```{r}
wb <- loadWorkbook("regional.xlsx")
i1 <- readWorksheet(wb, sheet=1, startRow=2)
cmix <- readWorksheet(wb, sheet=2, startRow=2)
c <- readWorksheet(wb, sheet=3, startRow=2)
ngmix <- readWorksheet(wb, sheet=4, startRow=2)
ng <- readWorksheet(wb, sheet=5,startRow=2)
pmix <- readWorksheet(wb, sheet=6, startRow=2)
p <- readWorksheet(wb, sheet=7, startRow=2)
#No mix data required. All 100% sourced at location.
nmix <- readWorksheet(wb, sheet=8, startRow=2)
hmix <- readWorksheet(wb, sheet=9, startRow=2)
gmix <- readWorksheet(wb, sheet=10, startRow=2)
smix <- readWorksheet(wb, sheet=11, startRow=2)
wmix <- readWorksheet(wb, sheet=12, startRow=2)
wbiomix <- readWorksheet(wb, sheet=13, startRow=2)

femix <- readWorksheet(wb, sheet=14, startRow=2)
fe <- readWorksheet(wb, sheet=15, startRow=2)

```

###Processing Sheet 1
```{r}
myseq <- seq(from=2, to=55, by=1)
names(i1) <- c('year',myseq)
ei1 <- melt(i1, id.vars="year")
ei1$variable <- as.character(ei1$variable)
ei1$comp_id <- 1
ei1$src_id <- 12

expi1 <- data.frame(cbind(ei1$year,ei1$variable,ei1$src_id,ei1$comp_id,ei1$value))
names(expi1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(expi1)){
  text <- paste(text, "insert into tbl_values values (",
                expi1[i,1],",",
                expi1[i,2],",",
                expi1[i,3],",",
                expi1[i,4],",",
                expi1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_index1.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 2
Source = 1 (Coal)
Component = 3 (Share of Energy Mix)
```{r}
names(cmix) <- c('year',myseq)
cmix1 <- melt(cmix, id.vars="year")
cmix1$variable <- as.character(cmix1$variable)
cmix1$comp_id <- 3
cmix1$src_id <- 1

cmix1 <- data.frame(cbind(cmix1$year,cmix1$variable,cmix1$src_id,cmix1$comp_id,cmix1$value))
names(cmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(cmix1)){
  text <- paste(text, "insert into tbl_values values (",
                cmix1[i,1],",",
                cmix1[i,2],",",
                cmix1[i,3],",",
                cmix1[i,4],",",
                cmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_coal_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 3
Source = 1 (Coal)
Component = 4 (percent sourced)
```{r}
names(c) <- c('year',myseq)
c1 <- melt(c, id.vars="year")
c1$variable <- as.character(c1$variable)
c1$comp_id <- 4
c1$src_id <- 1

c1 <- data.frame(cbind(c1$year,c1$variable,c1$src_id,c1$comp_id,c1$value))
names(c1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(c1)){
  text <- paste(text, "insert into tbl_values values (",
                c1[i,1],",",
                c1[i,2],",",
                c1[i,3],",",
                c1[i,4],",",
                c1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_coal.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 4
Source = 2 (Natural Gas)
Component = 3 (Share of Energy Mix)
```{r}
names(ngmix) <- c('year',myseq)
ngmix1 <- melt(ngmix, id.vars="year")
ngmix1$variable <- as.character(ngmix1$variable)
ngmix1$comp_id <- 3
ngmix1$src_id <- 2

ngmix1 <- data.frame(cbind(ngmix1$year,ngmix1$variable,ngmix1$src_id,ngmix1$comp_id,ngmix1$value))
names(ngmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(ngmix1)){
  text <- paste(text, "insert into tbl_values values (",
                ngmix1[i,1],",",
                ngmix1[i,2],",",
                ngmix1[i,3],",",
                ngmix1[i,4],",",
                ngmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_natural_gas_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```
###Processing Sheet 5
Source = 2 (Natural Gas)
Component = 4 (percent sourced)
```{r}
names(ng) <- c('year',myseq)
ng1 <- melt(ng, id.vars="year")
ng1$variable <- as.character(ng1$variable)
ng1$comp_id <- 4
ng1$src_id <- 2

ng1 <- data.frame(cbind(ng1$year,ng1$variable,ng1$src_id,ng1$comp_id,ng1$value))
names(ng1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(ng1)){
  text <- paste(text, "insert into tbl_values values (",
                ng1[i,1],",",
                ng1[i,2],",",
                ng1[i,3],",",
                ng1[i,4],",",
                ng1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_natural_gas.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 6
Source = 3 (Petroleum)
Component = 3 (Share of Energy Mix)
```{r}
names(pmix) <- c('year',myseq)
pmix1 <- melt(pmix, id.vars="year")
pmix1$variable <- as.character(pmix1$variable)
pmix1$comp_id <- 3
pmix1$src_id <- 3

pmix1 <-data.frame(cbind(pmix1$year,pmix1$variable,pmix1$src_id,pmix1$comp_id,pmix1$value))
names(pmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(pmix1)){
  text <- paste(text, "insert into tbl_values values (",
                pmix1[i,1],",",
                pmix1[i,2],",",
                pmix1[i,3],",",
                pmix1[i,4],",",
                pmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_petroleum_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 7
Source = 3 (Petroleum)
Component = 4 (percent sourced)
```{r}
names(p) <- c('year',myseq)
p1 <- melt(p, id.vars="year")
p1$variable <- as.character(p1$variable)
p1$comp_id <- 4
p1$src_id <- 3

p1 <- data.frame(cbind(p1$year,p1$variable,p1$src_id,p1$comp_id,p1$value))
names(p1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(p1)){
  text <- paste(text, "insert into tbl_values values (",
                p1[i,1],",",
                p1[i,2],",",
                p1[i,3],",",
                p1[i,4],",",
                p1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_petroleum.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 8
Sheets 8 - 13 only have mix information. In other words, these data 
are all 100% sourced domestically.
Source = 4 (Nuclear)
Component = 3 (Share of Energy Mix)
```{r}
names(nmix) <- c('year',myseq)
nmix1 <- melt(nmix, id.vars="year")
nmix1$variable <- as.character(nmix1$variable)
nmix1$comp_id <- 3
nmix1$src_id <- 4

nmix1 <-data.frame(cbind(nmix1$year,nmix1$variable,nmix1$src_id,nmix1$comp_id,nmix1$value))
names(nmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(nmix1)){
  text <- paste(text, "insert into tbl_values values (",
                nmix1[i,1],",",
                nmix1[i,2],",",
                nmix1[i,3],",",
                nmix1[i,4],",",
                nmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_nuclear_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 9
Source = 5 (Hydro)
Component = 3 (Share of Energy Mix)
```{r}
names(hmix) <- c('year',myseq)
hmix1 <- melt(hmix, id.vars="year")
hmix1$variable <- as.character(hmix1$variable)
hmix1$comp_id <- 3
hmix1$src_id <- 5

hmix1 <-data.frame(cbind(hmix1$year,hmix1$variable,hmix1$src_id,hmix1$comp_id,hmix1$value))
names(hmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(hmix1)){
  text <- paste(text, "insert into tbl_values values (",
                hmix1[i,1],",",
                hmix1[i,2],",",
                hmix1[i,3],",",
                hmix1[i,4],",",
                hmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_hydro_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 10
Source = 6 (Geothermal)
Component = 3 (Share of Energy Mix)
```{r}
names(gmix) <- c('year',myseq)
gmix1 <- melt(gmix, id.vars="year")
gmix1$variable <- as.character(gmix1$variable)
gmix1$comp_id <- 3
gmix1$src_id <- 6

gmix1 <-data.frame(cbind(gmix1$year,gmix1$variable,gmix1$src_id,gmix1$comp_id,gmix1$value))
names(gmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(gmix1)){
  text <- paste(text, "insert into tbl_values values (",
                gmix1[i,1],",",
                gmix1[i,2],",",
                gmix1[i,3],",",
                gmix1[i,4],",",
                gmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_geothermal_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 11
Source = 7 (Solar/PV)
Component = 3 (Share of Energy Mix)
```{r}
names(smix) <- c('year',myseq)
smix1 <- melt(smix, id.vars="year")
smix1$variable <- as.character(smix1$variable)
smix1$comp_id <- 3
smix1$src_id <- 7

smix1 <-data.frame(cbind(smix1$year,smix1$variable,smix1$src_id,smix1$comp_id,smix1$value))
names(smix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(smix1)){
  text <- paste(text, "insert into tbl_values values (",
                smix1[i,1],",",
                smix1[i,2],",",
                smix1[i,3],",",
                smix1[i,4],",",
                smix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_solarpv_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 12
Source = 8 (Wind)
Component = 3 (Share of Energy Mix)
```{r}
names(wmix) <- c('year',myseq)
wmix1 <- melt(wmix, id.vars="year")
wmix1$variable <- as.character(wmix1$variable)
wmix1$comp_id <- 3
wmix1$src_id <- 8

wmix1 <-data.frame(cbind(wmix1$year,wmix1$variable,wmix1$src_id,wmix1$comp_id,wmix1$value))
names(wmix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(wmix1)){
  text <- paste(text, "insert into tbl_values values (",
                wmix1[i,1],",",
                wmix1[i,2],",",
                wmix1[i,3],",",
                wmix1[i,4],",",
                wmix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_wind_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 13
Source = 9 (Wood/Waste)
Component = 3 (Share of Energy Mix)
```{r}
names(wbiomix) <- c('year',myseq)
wbiomix1 <- melt(wbiomix, id.vars="year")
wbiomix1$variable <- as.character(wbiomix1$variable)
wbiomix1$comp_id <- 3
wbiomix1$src_id <- 9

wbiomix1 <-data.frame(cbind(wbiomix1$year,wbiomix1$variable,wbiomix1$src_id,wbiomix1$comp_id,wbiomix1$value))
names(wbiomix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(wbiomix1)){
  text <- paste(text, "insert into tbl_values values (",
                wbiomix1[i,1],",",
                wbiomix1[i,2],",",
                wbiomix1[i,3],",",
                wbiomix1[i,4],",",
                wbiomix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_wood_bio_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 14
Source = 10 (Fuel Ethanol)
Component = 3 (Share of Energy Mix)
```{r}
names(femix) <- c('year',myseq)
femix1 <- melt(femix, id.vars="year")
femix1$variable <- as.character(femix1$variable)
femix1$comp_id <- 3
femix1$src_id <- 10

femix1 <-data.frame(cbind(femix1$year,femix1$variable,femix1$src_id,femix1$comp_id,femix1$value))
names(femix1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(femix1)){
  text <- paste(text, "insert into tbl_values values (",
                femix1[i,1],",",
                femix1[i,2],",",
                femix1[i,3],",",
                femix1[i,4],",",
                femix1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_fuel_ethanol_mix.sql")
writeLines(text, fileConn)
close(fileConn)
```

###Processing Sheet 15
Source = 10 (Fuel Ethanol)
Component = 4 (Share of Energy Mix)
```{r}
names(fe) <- c('year',myseq)
fe1 <- melt(fe, id.vars="year")
fe1$variable <- as.character(fe1$variable)
fe1$comp_id <- 4
fe1$src_id <- 10

fe1 <-data.frame(cbind(fe1$year,fe1$variable,fe1$src_id,fe1$comp_id,fe1$value))
names(fe1) <- c('year_id','reg_id','src_id','comp_id','measurement')

text <- NULL
for(i in 1:nrow(fe1)){
  text <- paste(text, "insert into tbl_values values (",
                fe1[i,1],",",
                fe1[i,2],",",
                fe1[i,3],",",
                fe1[i,4],",",
                fe1[i,5],");\n",sep="")
}
fileConn<-file("export_regions_fuel_ethanol.sql")
writeLines(text, fileConn)
close(fileConn)
```